---
title: "Class 12: Lab  and HW Q13/14"
author: "Jervic Aquino (PID:A17756721)"
date: 02-12-2026
toc: TRUE
format: pdf
---

## Lab Section 1 Work

Download then read the CSV file from Ensemble


```{r}
data <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(data)
```

```{r}
table(data$Genotype..forward.strand.)
```

```{r}
table(data$Genotype..forward.strand.) / nrow(data)
```

## Section 4 HW

How many samples do we have ?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```


> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

- Sample size: 

```{r}
table(expr$geno)
```


```{r}
median <- boxplot(exp ~ geno, expr)

print(median)
```

- Corresponding median expression levels:

```{r}
print(paste(median$names, median$stats[3,]))
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr) + 
  aes(x=geno, y=exp) + 
  geom_boxplot()
```

- It can inferred that the A/A genotype causes higher levels of expression than the G/G genotype. The SNP effect does effect the expression of ORMDL3, in which ORMDL3 has an expression level in between A/A and G/G. This may be because having an A allows for expression higher than G/G but being accompanied by G results in expression less than A/A. 
