---
title: "Class 5: Data Vis with ggplot"
author: "Jervic Aquino 
(PID:A17756721)"
format: pdf
toc: TRUE 
---

## Background

There are lot's of ways to make plots in R. These include so-called "base R" (like the `plot()`) and add on packages like **ggplot2**.

Let's make the same plot with these two graphics systems. We can use the inbuilt `cars` dataset:

```{r}
head(cars)
```

With "base R" we can simply: 

```{r}
plot(cars)
```

Now let's try ggplot. First I need to install the package using `install.packages("ggplot2")`. 

> **N.B.** We never run an `install.packages()` in a code chunk otherwise we will re-install needlessly every time we render our document. 

Everytime we want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least three things:

1. The **data** i.e. stuff to plot as a data.frame
2. The **aes** or aesthetics that map the data to the plot 
3. The **geom_** or geometry i.e. the plot type such as points, lines, etc. 


```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + geom_line() 
```
```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth() 
```

```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x="Speed (MPH)", 
       y = "Distance (ft)", 
       title = "Stopping Distance of Old Cars") + theme_bw() 
```







## Gene Expression Plot 

Read some data on the effects of GLP-1 inhibitor (drug) on gene expression values. 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
Version 1 Plot - start simple by getting some ink on the page.

```{r}
ggplot(genes) + aes(Condition1, Condition2) + geom_point(col="blue", alpha=0.2)
```

Let's color by `State` up, down or no change. 

```{r}
table(genes$State) 
```

```{r}
ggplot(genes) + aes(Condition1, Condition2, col=State) + geom_point() +
  scale_color_manual(values = c("purple", "gray", "orange")) + 
  labs(x="Control (no drugs)", 
       y= "Drug", 
       title = "Expression Changes with GLP-1 Drug") + theme_bw()
```

## Going further with gapmider

Here we explore the famous `gapminder` dataset with some custom plots. 


```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many rows does this dataset have?

```{r}
nrow(gapminder)
```

> Q. How many different continents are in this dataset ?

```{r}
table(gapminder$continent)
```



Version 1 plot GDP vs LifeExp for all rows
```{r}
ggplot(gapminder) + aes(gdpPercap, lifeExp, col=continent) + geom_point() + 
  labs(x="GDP per Capita", y="Life Expectancy") + theme_bw() 
```


I want to see a plot for each continent - in ggplot lingo this is called "Faceting"
```{r}
ggplot(gapminder) + aes(gdpPercap, lifeExp, col=continent) + geom_point() + 
  labs(x="GDP per Capita", y="Life Expectancy") + 
  theme_bw() + facet_wrap(~continent) 
```


## First look at the dplyr package

Another add-on package with a function called `filter()` that we want to use. 

```{r}
library(dplyr)
```


```{r}
filter(gapminder, year == 2007, country == "Ireland")
```


```{r}
filter(gapminder, year == 2007, country == "United States")
```



```{r}
input <- filter(gapminder, year == 2007 | year == 1977)

ggplot(input) + aes(gdpPercap, lifeExp, col=continent) +
  geom_point() + facet_wrap(~year) 
```

