---
title: "Class06: R Functions"
author: "Jervic Aquino (PID:A17756721)"
date: "1/23/2026"
format: pdf
toc: TRUE
---

## Background

Function are at the heart of using R. Everything we do involves calling using functions (from data input, analysis to results output). 

All functions in R have at least three things:

1. A **name**: the thing we use to call the function
2. One or more input **arguments** that are comma separated
3. The **body**: lines of code between curly brackets { } that does the work of the function

## A first function


Let's write a function that adds some numbers:
```{r}
add <- function(x) {
  x + 1
}
```


Let's try it out:
```{r}
add(100)
```


Will this work?
```{r}
add(c(100, 200, 300))
```


Modify to be more useful and add more than just 1
```{r}
add <- function(x, y=1) {
  x + y
}
```


```{r}
add(100, 10)
```


Will this still work?
```{r}
add(100)
```
No, because there is no y-component. We have to set y equal to value if adding only one argument or component in the created function. 


How to argue: 
```{r}
plot(1:10, col="blue", typ="b")
```

```{r}
log(10, base=10)
```


> **N.B.** Input arguments can be either **required** or **optional**. The latter have a fall-back default that is specified in the function code with an equals sign. 


```{r}
#add(100, 200, 300)
```


## A second function

All functions in R look like this

```
name <- function(arg) {
  body
}
```

The `sample()` function in R generates random numbers from the range of numbers given. It randomly picks items from the vector inputted. 

```{r}
sample(1:10, size=4)
```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, size=12, replace=TRUE)
```


> Q. Write the code to generate a random 12 nucleotide long DNA sequence 

```{r}
sample(c("a", "c", "g", "t"), size=12, replace=TRUE)
```

Another way to write the code:
```{r}
bases <- c("A", "T", "G", "C")
sample(bases, size=12, replace=TRUE)
```


> Q. Write a first version function called `generate_dna()` that generates a user specified length `n`, random DNA sequence

```{r}
n <- sample(c(4:30), size=1)
generate_dna <- sample(bases, size = n, replace = TRUE)
generate_dna
```

```{r}
generate_dna <- function(n=6) {
  bases <- c("A", "T", "G", "C") 
  sample(bases, size=n, replace=TRUE)
}
```

```{r}
generate_dna(100)
```


> Q. Modify your function to return a FASTA-like sequence so rather than [10] "G" "C" "A" "A" "T" we want "GCAAT" 

```{r}
generate_dna <- function(n=6) {
  bases <- c("A", "T", "G", "C") 
  sequence <- sample(bases, size=n, replace=TRUE)
  sequence <- paste(sequence, collapse="")
  return(sequence)
}
```

```{r}
generate_dna(10)
```


> Q. Give the user an option to return FASTA format output sequence or standard multi-element vector format

```{r}
generate_dna <- function(n=6, fasta=TRUE) {
  bases <- c("A", "T", "G", "C") 
  sequence <- sample(bases, size=n, replace=TRUE)
  
  if(fasta) {
  sequence <- paste(sequence, collapse="")
  cat("Hello...")
  } else {
    cat("is it me you're looking for...")
  }
  
  return(sequence)
}  
```

```{r}
generate_dna(10)
```
```{r}
generate_dna(10, fasta = FALSE)
```



## A new cool function

> Q. Write a function called `generate_protein()` that generates a user specific length protein sequence in FASTA format

```{r}
generate_protein <- function(n) {
  aa <- sample(c("A","R","N","D","C","Q","E","G","H",
        "I","L","K","M","F","P","S","T","W","Y","V"), size=n, replace=TRUE)
  protein <- paste(aa, collapse="")
  return(protein)
}
```

```{r}
generate_protein(10)
```


> Q. Use your new `generate_protein()` function to generate sequences between length 6 and 12 amino acids in length and check if any of these are unique in nature (i.e. found in the NR database at NCBI)

```{r}
generate_protein <- function(n) {
  aa <- sample(c("A","R","N","D","C","Q","E","G","H",
        "I","L","K","M","F","P","S","T","W","Y","V"), size=n, replace=TRUE)
  protein <- paste(aa, collapse="")
  return(protein)
}
```

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
```

Or we could do a `for()` loop:

```{r}
for(i in 6:12) {
  cat(">", i, sep="", "\n" )
  cat(generate_protein(i), "\n")
}
```


